# FinalProjectNimra

### Multi-Path AIC Model Selection with Stability and Plausibility Filtering

This R package implements a **multi-path model selection algorithm** based on the Akaike Information Criterion (AIC). It automatically explores multiple near-optimal model paths, evaluates **variable stability** through resampling, and identifies **plausible, stable models** for both linear and logistic regression tasks.

---

## Installation

You can install the package directly from GitHub using:

```{r}
# Install devtools if not already available
install.packages("remotes")

# Install the package
remotes::install_github("R-4-Data-Science/FinalProjectNimra")

#Then load it with:
library(FinalProjectNimra)

```

## Example 1 – Linear Regression

Below is a minimal example demonstrating the end-to-end workflow on the built-in mtcars dataset.

```{r}
set.seed(123)

# Run full multi-path model selection
res_linear <- build_paths(
  data = mtcars,
  response = "mpg",
  K = 4,       # number of forward steps
  B = 20,      # bootstrap resamples
  delta = 1,   # AIC tolerance
  tau = 0.6    # stability threshold
)

# List plausible models
res_linear$plausible_models

# Plot variable stability
plot_stability(res_linear$stability)
```

## Example 2 – Logistic Regression

This example uses the same data but treats am (automatic/manual transmission) as a binary response.

```{r}
set.seed(123)

# Run logistic model selection
res_logistic <- multi_path_AIC_procedure(
  data = mtcars,
  response = "am",
  K = 4,
  B = 20,
  delta = 1,
  tau = 0.6
)

# Display plausible logistic models
res_logistic$plausible_models

# Plot stability
plot_stability(res_logistic$stability)

```

## Summary

The package demonstrates how multi-path AIC selection and stability analysis together provide
a robust framework for identifying reproducible, interpretable models in both linear and logistic regression.
